Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-09-17 doc/response_to_reviewer_R1.Rmd

•	276: This section is important, but it's not a relevant result; rather, it's a methodological aspect that allowed the study to be conducted. I think it should be included in the materials and methods section.

**RESPONSE: Thanks for this suggestion. We have now moved the description of the droughts to the methods section and have only retained a brief description of the differences among droughts in the results.**

•	Since the study is based on only one site, it is relevant to improve the discussion section with other already published study sites (add more references), mainly link the global implications, such mortality rate and vulnerability in tropical forests. See https://urldefense.proofpoint.com/v2/url?u=https-3A__doi.org_10.1126_science.adq6607&d=DwIFaQ&c=009klHSCxuh5AI1vNQzSO0KGjl4nbi2Q0M1QLJX9BeE&r=6fH468T58at0TtM59sMf109A_dRgo6XTfGMOVX5Ewrw&m=fPE_4y1UPsTl9St_Y44IUO7VIaQJOJIUiHtK15-Nr7g4UiVKbtYQ-TE0043SLAj6&s=SxqVZgFkbqAZ0-VEvdQntXWpGEUu9lr25rgxQtwaVdw&e=

*RESPONSE: Thanks for this suggestion. We were excited to see this global study published recently. We have now discussed our focused work in relation to this global study based on the alignment of the a key message - the diversity of drought responses with drought types and species traits (deciduousness). However, dendroband datasets can be key to filling gaps in spatial and climatic coverage of tree rings records in the tropics (reviewed by @groenendijk_importance_2025). Moreover, our results showcasing response diversity among individuals within a species demonstrates an important caveat to aggregate/species-chronology-based inferences from tree-ring records.*

@@

Figure 1a: “Dy” is not correct

**RESPONSE: Thanks for noticing this error. We have fixed this to say "Day of year".**

Fig. 2. It is unnecessary to include the years 2023 and 2024 on the y-axis, as they were not evaluated. I think this figure could be analyzed further. For example, it took two years after 2014 for growth to recover to previous year (for ALL species) unlike in 2010 and 2020, when it rebounded within a year to levels equal to or higher than the previous year.

*RESPONSE: Thanks for this suggestion. We have removed 2023 and 2024 from the x-axis. While we agree that there is a lot more to discuss and analyse from this figure, we wanted to be careful not to overinterpret the growth patterns without adequate analysis of other covariates. For example, the pattern of suppressed growth after the 2015 drought is not necessarily because of lag times in recovery but more likely because drought conditions extended into 2016 - this was a multi-year drought - and El Nino conditions switched to La Nina only late in 2016, leading to a late monsoon. Our analysis was focussed on episodic droughts, and therefore we chose not to analyse and interpret the whole timeseries in this study. We recognise however, that lagged responses and compensatory responses would be important in such timeseries and hope to answer these as the dataset accumulates.*

### 2025-09-17 doc/manuscript_hkk_drought_sensitivity.Rmd
@teixeirak full draft of responses
- tentative answers in italics
- commit includes track changes version with _track in name

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 4999   
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

>>>>

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 5074   
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022, @groenendijk_importance_2025], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->
